---
title: "Parameters and literature references"
author: "George G. Vega Yon, Ph.D."
date: today
vignette: >
  %\VignetteIndexEntry{Parameters and literature references}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
references:
- id: guerra2017
  type: article-journal
  title: "The basic reproduction number (R0) of measles: a systematic review"
  author:
  - family: Guerra
    given: Fiona M.
  - literal: "others"
  container-title: The Lancet Infectious Diseases
  volume: 17
  issue: 12
  page: e420-e428
  issued:
    date-parts:
    - - 2017
  DOI: 10.1016/S1473-3099(17)30307-9
- id: diekmann2010
  type: article-journal
  title: "The construction of next-generation matrices for compartmental epidemic models"
  author:
  - family: Diekmann
    given: O.
  - family: Heesterbeek
    given: J. A. P.
  - family: Roberts
    given: M. G.
  container-title: Journal of The Royal Society Interface
  volume: 7
  issue: 47
  page: 873-885
  issued:
    date-parts:
    - - 2010
  DOI: 10.1098/rsif.2009.0386
- id: jones2026
  type: article-journal
  title: "Measles Hospitalizations in Utah, 2025–2026"
  author:
  - family: Jones
    given: Anna
  - literal: "others"
  container-title: NEJM Evidence
  volume: 5
  issue: 8
  issued:
    date-parts:
    - - 2026
  DOI: 10.1056/EVIDpha2600141
- id: wang2026
  type: article-journal
  title: "Characteristics of Patients Hospitalized with Measles During an Outbreak — West Texas, January–March 2025"
  author:
  - family: Wang
    given: Dennis
  - literal: "others"
  container-title: MMWR. Morbidity and Mortality Weekly Report
  volume: 75
  issue: 20
  page: 252-257
  issued:
    date-parts:
    - - 2026
  DOI: 10.15585/mmwr.mm7520a1
- id: utahdhhs2026
  type: report
  title: "Measles Disease Plan"
  author:
  - literal: "Utah Department of Health and Human Services"
  issued:
    date-parts:
    - - 2026
      - 3
      - 26
  URL: https://epi.utah.gov/wp-content/uploads/Measles-disease-plan.pdf
---

<!-- how-to-cite -->
::: {.callout-note}
## How to cite

If you use **measles** in published work, please cite it — run
`citation("measles")` in R for the full entry.
:::
<!-- how-to-cite -->

This vignette documents the parameters of the measles models in this package: their default values, where those values come from, and how far each has been verified. The same information is available as a data frame through `measles_parameters()`, which reads the file `inst/extdata/measles_parameters.csv`. That file is the canonical parameter table for our measles projects; projects that use different values should list them next to the package defaults and explain why.

```{r}
#| label: setup
library(measles)
params <- measles_parameters()
```

## The parameter table

Each row of the table is a parameter, with its default value, units, and source. The current edition of the Utah DHHS Measles Disease Plan cited here was last updated on March 26, 2026 (footer dated February 2026) [@utahdhhs2026].

```{r}
#| label: params-table
#| echo: false
#| output: asis
tab <- format(
  params[, c(
    "parameter", "r_argument", "default", "units", "citation",
    "doi_or_url", "notes"
  )],
  markdown = TRUE
)

cat(
  "| Parameter (argument) | Default (units) | Source | Notes |",
  "|---|---|---|---|",
  paste0("| ", do.call(paste, c(tab, sep = " | ")), " |"),
  sep = "\n"
)
```

The `models` column lists the functions that take each argument. For instance, the parameters of `ModelMeaslesSchool()` are:

```{r}
#| label: school-params
measles_parameters("ModelMeaslesSchool")[, c("r_argument", "default", "units")]
```

`InterventionMeaslesPEP()` has no default values; the table lists the values used in `vignette("school")`.

## Calibrating to R0

Measles is highly transmissible, with a basic reproduction number (R0) usually reported between 12 and 18 [@guerra2017]. Our models target R0 = 15, the midpoint of that range, which implies a herd-immunity threshold of $1 - 1/R_0 \approx 93\%$.

Instead of estimating both the transmission probability and the contact rate, we fix the transmission probability at a high value (0.9, consistent with the Utah DHHS plan statement that 90% of susceptible contacts develop disease) and calibrate the number of contacts to match the target R0.

In `ModelMeaslesSchool()`, agents are infectious during the prodromal period only (detected cases with rash are isolated), so

$$
R_0 = p_t \times c \times D_p,
$$

where $p_t$ is `transmission_rate`, $c$ is `contact_rate`, and $D_p$ is `prodromal_period`. The default contact rate solves this equation for R0 = 15:

```{r}
#| label: school-contact-rate
15 / 0.9 / 4
```

In the mixing models, the contact matrix holds the expected number of contacts between groups. The next-generation-matrix method [@diekmann2010] gives R0 as the spectral radius of the matrix of expected secondary infections. `calibrate_mixing_model()` finds the factor that scales the contact matrix so it matches the target R0. When agents with rash are not fully isolated (`rash_reduction_contact_rate < 1` in `ModelMeaslesMixing()`), the infectious period used for calibration is $D_p + (1 - r_r) D_r$, where $D_r$ is the rash period and $r_r$ is the reduction in contacts during rash.

## Hospitalization: rate vs probability

The models take a daily hospitalization **rate**, not a probability. During the rash period, an agent competes between hospitalization (rate $h$) and recovery (rate $1/D_r$), so the probability of being hospitalized is

$$
P(\text{hospitalization}) = \frac{h}{h + 1/D_r}.
$$

With the defaults, $h = 0.2$ and $D_r = 3$ days, this probability is:

```{r}
#| label: hosp-prob
h <- 0.2
rash_period <- 3
h / (h + 1 / rash_period)
```

To match a target probability $p$, invert the formula, $h = p \times (1/D_r) / (1 - p)$. For example, recent analyses use $p = 0.1$, which is consistent with the 8% overall (9% among the unvaccinated) reported in Utah [@jones2026]:

```{r}
#| label: hosp-rate
p_hosp <- 0.1
p_hosp * (1 / rash_period) / (1 - p_hosp)
```

The default of 0.2 started as a conservative value agreed with Utah DHHS. For comparison, 18.5% of patients were hospitalized during the 2025 outbreak in West Texas [@wang2026].

## Inactive parameters

The argument `vax_reduction_recovery_rate` of `ModelMeaslesMixing()` is not active. The C++ model registers it as "(IGNORED) Vax improved recovery", and it has no effect on the simulations. `ModelMeaslesSchool()` no longer takes it.

## References

::: {#refs}
:::
